Floral selection in mountain bumble bees and a test of the 'target use hypothesis

### 2021-06-16 ms_2/ms_2.Rmd
finalized new draft

---
title: "Floral selectivity in mountain bumble bees and a test of the 'target use hypothesis'"

@@

Understanding patterns of resource use by animals, and the underlying selection processes that govern these patterns, can clarify how ecological communities have assembled, how they currently function, and how they are likely to respond to future perturbations. Communities of bumble bees and their floral hosts provide a tractable and compelling system in which to study questions of resource use, and the answers to these questions have immediate relevance to the conservation of bumble bees and their mutualists under the forcings of climate and land use change, particularly in mountain ecosystems.

Using a dataset consisting of ~13,000 observations collected along a 1400 m elevation gradient over three consecutive years, we

Significance ...

*Keywords*: foraging, trait matching, pollinator, *Bombus*, alpine

@@

@@

Given the well-established mechanistic link between bumble bee tongue length and floral selection [e.g. @Brian1954-cy; @Heinrich1976-wc; @Teras1976-pi; @Inouye1978-qi; @Pyke1982-gn], our finding that bumble bee tongue length explained selectivity (i.e. null model residuals) much better than it explained raw visitation rates indicates that, in our study system, abundance and preference function jointly in bumble bee floral choice. This casts doubt on the *target use hypothesis* as a general rule of bumble bee foraging. It does not, however, mean that the *target use hypothesis* does not obtain under certain circumstances, such as when a highly preferred resource varies within a nonlimiting rage of availability. Moreover, our results suggest something akin to the *target use hypothesis* but formulated negatively with respect to resources that are *not selected*. Visitation rates to disc flowers in our system, for example, were very low despite the fact that they often occurred at high abundance, suggesting that while preference seems to interact with resource abundance, avoidance may not. In general, though, our findings support the hypothesis that bumble bee resource selection involves both the neutral effects of abundance and the directed effect of preference, which implies that analytical techniques designed to parse these effects, such as the null model analysis we employ, should be favored in studies of bumble bee foraging. 
<!-- Accordingly, we shall focus primarily on our calculations of selectivity, rather than raw visitation rates, as we interpret the other results of our study. -->

@@

*First, talk about the lability of selectivity, interesting difference between species, caveats.*

General patterns

While the floral choices of bumble bees have been studied extensively in both laboratory [e.g. @Kugler1943-uy; @Raine2007-nq; @Maharaj2019-wm] and field studies [e.g. @Heinrich1976-wc; @Teras1976-pi; @Goulson2008-zy], studies of the *dynamics* of bumble bee floral choice along ecological gradients are rare [e.g. @Miller-Struttmann2014-wz], and we are not aware of any that use null model controls to distinguish selectivity from the neutral effects of relative abundance.

Interesting deviations

The late high elevation peak of visitation and selectivity between soro and bell flowers is probably driven by soro's peculiar affinity for Campanula.

We reiterate, however, the inherent problems of comparing selectivity --- understood as an index of the latent process of preference --- across different contexts of resource abundance. We recommend the cautious interpretation of selectivity patterns through elevation and time in conjunction with the underlying patterns of floral abundance and raw visitation.

*Then, segue to climate and land use.*

Biological patterns through elevation and time are acted upon by climate and land use change, and the consequence of this action is the generation of new biological patterns through elevation and time. A central task of ecology is to understand the patterns that will be acted upon and predict the ecological ramifications of the patterns that will be created. In this regard, it is significant that we find marked structure in the morphotypal composition. raw visitation, and selectivity of the floral community on which the bumble bees of the Berchtesgadener Alps depend. While our data permit only speculation as to the climate and land use driven processes at work in our system, our speculation can be nonetheless well-informed. Climate change over the last century has caused significant upslope movement [e.g. @Marshall2020-pc] and phenological shifts [e.g. @CaraDonna2014-nq] of floral resources for pollinators. It is likely that the current patterns of morphotype relative abundance that we observed differ from those that would have been observed a century ago, and in the coming decades warming climates will probably induce further shifts upslope [@Marshall2020-pc] and potentially complex shifts in the onset, peak, and end of flowering periods [@CaraDonna2014-nq]. These effects of climate change will interact with the likely more rapid effects of land use change. In particular, the reforestation of montane meadows can be expected to erode the abundance of flowers from mid-elevation up to the tree line (in our system, around ~1500 m.s.l.). Above the tree line, however, floral communities should be relatively safe from reforestation since the advance of alpine tree lines tends to lag behind climate change [@Dullinger2004-qu]. At lower elevations, agricultural intensification can be expected to suppress overall floral availability by favoring grasses over forbs and shift the composition of the community toward species that thrive under the altered soil conditions and disturbance regimes of intensive agriculture [@Tasser2002-om].

## 4.4 Final caveats

In addition to the qualifying remarks provided in the sections above, it is important to acknowledge some limitations of our study that should caution the reader against the over-interpretation of our results. First, the inference of selection as use-relative-to-abundance is sensitive to exactly *how* abundance is estimated [@Johnson1980-co; @Beyer2010-vr]. Of particular concern in our study is the problem of spatial scale. By necessity, we measured floral abundance and visitation along transects *within* our study sites. Bumble bee foraging range, however, is much larger than our study sites, and it is possible that the relative abundances of floral morphotypes outside our study sites but within foraging range differ from our within-site estimates. Moreover, because bumble bees can forage across multiple meadow patches in a mountain forest matrix [@Mola2020-yc], there exists a prior selective process --- Johnson's (1980) "second-order selection" --- in which bumble bees choose among available foraging areas before selecting from resources within an area. Finally, to account for the neutral effects of abundance on floral use, we used a simple null model in which the probability of a resource being selected was assumed to be linearly proportional to the relative abundance of the resource. It is possible, however, for resource use to exhibit nonlinear "functional responses" @Mysterud1998-rs to resource abundance, which, relative to a linear null model, would give biased estimates of selectivity. Indeed, the fact that patterns of selectivity through elevation and time **(Figure 8)** partly retain of the pattern of resource abundance **(Figure 7)** suggests that there may be some nonlinear dependence of usage on abundance that cannot fully be removed by our linear null model.

# 5. CONCLUSIONS

Understanding patterns of resource use by animals, and the underlying selection processes that govern these patterns, can clarify how ecological communities have assembled in the past, how they currently function, and how they are likely to respond to future perturbations. Communities of bumble bees and their floral hosts provide a tractable and compelling system in which to study questions of resource use, and the answers to these questions have immediate relevance to the conservation of bumble bees and their mutualists under the forcings of climate and land use change, particularly in mountain ecosystems.

<!-- Resource usage is typically studied as a phenomenological level, with questions of mechanism relegated to speculations regarding the latent process of resource selection that lurks behind realized use patterns [@Beyer2010-vr]. In our study, we evaluate the target use hypothesis by reference to a known mechanism of floral selection, namely bumble bee tongue length. -->

>>>>

---
title: "Floral selection in mountain bumble bees and a test of the 'target use hypothesis'"

@@

Understanding patterns of resource use by animals, and the underlying selection processes that govern these patterns, can clarify how ecological communities have assembled in the past, how they currently function, and how they are likely to respond to future perturbations. Communities of bumble bees and their floral hosts provide a tractable and compelling system in which to study questions of resource use, and the answers to these questions have immediate relevance to the conservation of bumble bees and their mutualists under the forcings of climate and land use change, particularly in mountain ecosystems. Using a dataset consisting of ~13,000 interactions recorded along a 1400 m elevation gradient over three consecutive years, we analyze bumble bee resource selection in terms of floral morphotype. We find that most bumble bee species exhibit strong positive selection toward flowers of the bell, lip, and/or flag morphotypes. Species-level preferences for floral morphotypes were explained well by bumble bee tongue length, which is well-established a key matching trait that constrains bumble bee floral choice. Tongue length was a much stronger predictor of null-model-corrected visitation rates than of raw visitation rates, indicating that, contrary to the *target use hypothesis*, neutral effects of resource abundance function in conjunction with learned or innate floral biases to determine bumble bee floral choice. By modeling floral selection through elevation and time, we demonstrate strong dynamics of resource availability and resource selection, suggesting that bumble bees have evolved flexible morphotype preferences suited to the extreme spatiotemporal floristic turnover of mountain ecosystems. Patterns of resource availability and selection through elevation and time are likely to be distorted in coming decades by the simultaneous processes of climate and land use change, which together cause the displacement and redistribution of mountain floral communities.

*Keywords*: foraging, trait matching, pollinator, *Bombus*, alpine, niche, resource selection

@@

In addition to the qualifying remarks above, it is important to acknowledge some additional limitations of our methods. First, the inference of selection as use-relative-to-abundance is sensitive to exactly *how* abundance is estimated [@Johnson1980-co; @Beyer2010-vr]. Of particular concern in our study is the problem of spatial scale. By necessity, we measured floral abundance and visitation along transects *within* our study sites. Bumble bee foraging range, however, is much larger than our study sites, and it is possible that the relative abundances of floral morphotypes outside our study sites but within foraging range differ from our within-site estimates. Moreover, because bumble bees can forage across multiple meadow patches in a mountain forest matrix [@Mola2020-yc], there exists a prior selective process --- Johnson's (1980) "second-order selection" --- in which bumble bees choose among available foraging areas before selecting from resources within an area. Finally, to account for the neutral effects of abundance on floral use, we used a simple null model in which the probability of a resource being selected was assumed to be linearly proportional to the relative abundance of the resource. It is possible, however, for resource use to exhibit nonlinear "functional responses" @Mysterud1998-rs to resource abundance, which, relative to a linear null model, would give biased estimates of selectivity.

@@

Given the well-established mechanistic link between bumble bee tongue length and floral selection [e.g. @Brian1954-cy; @Heinrich1976-wc; @Teras1976-pi; @Inouye1978-qi; @Pyke1982-gn], our finding that bumble bee tongue length explained selectivity (i.e. null model residuals) much better than it explained raw visitation rates indicates that, in our study system, abundance and preference function jointly in bumble bee floral choice. This casts doubt on the *target use hypothesis* as a general rule of bumble bee foraging. It does not, however, mean that the *target use hypothesis* does not obtain under certain circumstances, such as when a highly preferred resource varies within a nonlimiting rage of availability. Moreover, our results suggest something akin to the *target use hypothesis* but formulated negatively with respect to resources that are *not selected*. Visitation rates to disc flowers in our system, for example, were very low despite the fact that they often occurred at high abundance, suggesting that while preference seems to interact with resource abundance, avoidance may not. In general, though, our findings support the hypothesis that bumble bee resource selection involves both the neutral effects of abundance and the directed effect of preference, which implies that analytical techniques designed to disentangle these effects, such as the null model analysis we employ, should be favored in studies of bumble bee foraging. 
<!-- Accordingly, we shall focus primarily on our calculations of selectivity, rather than raw visitation rates, as we interpret the other results of our study. -->

@@

The intraspecific lability of bumble bee foraging preferences (*sensu stricto*) has been demonstrated under laboratory conditions [@Ings2009-vc; @Maharaj2019-wm]; our study reveals the *in situ* lability of selectivity for multiple bumble bee species through elevational and temporal gradients of floristic turnover. While we stress the limitations of interpreting empirical selectivity in terms of behavioral preference without controlling the joint abundances of all available resources [@Beyer2010-vr], interpreting the dynamics of selectivity in our study as the adaptive tuning of preference is consistent with the ecological hurdles bumble bees face as characteristically --- and most likely primordially --- mountain species [@Williams1985-ql; @Hines2008-er]. Where steep temperature gradients and short growing seasons generate extreme spatiotemporal turnover of floral resources, adaptive flexibility of morphological preferences should be favored over fixed innate biases.

At least three mechanisms could underlie the patterns of selectivity through elevation and time that we observed. To illustrate these hypotheses, we will take an as example the relatively strong selectivity dynamics of flag flowers in our system, which tended to be most preferred at high-elevation sites **(Figure 8)**. Our study does not enable the testing of these hypothesized mechanisms (and their constituent submechanisms), but we articulate them for the sake of clarity and as a guide for future investigations. Moreover, the hypotheses we propose not mutually exclusive, and it plausible that all three could obtain simultaneously in a given ecological context.

First, variation in selectivity toward a given morphotype could be driven by changes in the *intrinsic* reward of that focal morphotype. These changes might, in turn, be driven by (1) shifts in species composition of the focal morphotype (e.g. the occurrence of a particularly rewarding flag flower at high-elevation sites), (2) shifts in the reward levels of individual species (e.g. increased nectar production in the flag flower *Lotus corniculatus* under high elevation conditions), or (3) possibly changes in nutritional requirements of bumble bees (e.g. an increased need for concentrated nectar to support thermoregulation at colder ambient temperatures). The hypothesis is consistent with the interpretation of selectivity as an index of the latent process of absolute preference.

Second, selectivity toward a given morphotype could vary due to changes in the *relative* reward of the focal morphotype under changing background resource conditions. For example, it could be that flag flowers are selected more positively at high elevation because they must compete for visitation with highly attractive lip flowers at lower elevation. This hypothesis expresses the fundamental problem of inferring preference *sensu stricto* from selection, namely contingency of selection upon the joint availability of all resource options [@Johnson1980-co; Aarts2008-fo; @Beyer2010-vr].

The two hypotheses above are both grounded in the intrinsic properties of the floral community, whether those of the focal resource of those of the background resources. It is also possible, however, that patterns of resource selection could vary due to competitive interactions within or between bumble bee species. For example, the increased selection of flag flowers at high elevation could be driven by competition release from *B. pascuorum*, which tends to be the most abundance bumble bee species at lower elevations but becomes relatively scarce above 1500 m (Sponsler et al, submitted). Under this hypothesis, an organism's resource selection is the product of exogenous constraint rather than endogenous preference. For the reasons discussed above, however, we consider it unlikely that competition plays a constitutive role in governing bumble bee resource selection in the Berchtesgadener Alps.

<!-- Patterns of selectivity through elevation and time toward major floral morphotypes tended to be similar across the six dominant bumble bee species we  -->

<!-- Interesting deviations -->

<!-- The late high elevation peak of visitation and selectivity between soro and bell flowers is probably driven by soro's peculiar affinity for Campanula.  -->

<!-- Something about the apparent link between overall selectivity and its variability? -->

<!-- <!-- We reiterate, however, the inherent problems of comparing selectivity --- understood as an index of the latent process of preference --- across different contexts of resource abundance. We recommend the cautious interpretation of selectivity patterns through elevation and time in conjunction with the underlying patterns of floral abundance and raw visitation. --> -->

Biological patterns through elevation and time are acted upon by climate and land use change, and the consequence of this action is the generation of new biological patterns through elevation and time. A central task of ecology is to understand the patterns that will be acted upon and predict the ecological ramifications of the patterns that will be created. In this regard, it is significant that we find marked structure in the morphotypal composition, raw visitation, and selectivity of the floral community on which the bumble bees of the Berchtesgadener Alps depend. While our data permit only speculation as to the climate and land use driven processes at work in our system, our speculation can be nonetheless well-informed. Climate change over the last century has caused significant upslope movement [e.g. @Marshall2020-pc] and phenological shifts [e.g. @CaraDonna2014-nq] of floral resources for pollinators. It is likely that the current patterns of morphotype relative abundance that we observed differ from those that would have been observed a century ago, and in the coming decades warming climates will probably induce further shifts upslope [@Marshall2020-pc] and potentially complex shifts in the onset, peak, and end of flowering periods [@CaraDonna2014-nq]. These effects of climate change will interact with the likely more rapid effects of land use change. In particular, the reforestation of montane meadows can be expected to erode the abundance of flowers from mid-elevation up to the tree line (in our system, around ~1500 m.s.l.). Above the tree line, however, floral communities should be relatively safe from reforestation since the advance of alpine tree lines tends to lag behind climate change [@Dullinger2004-qu]. At lower elevations, agricultural intensification can be expected to suppress overall floral availability by favoring grasses over forbs and shift the composition of the community toward species that thrive under the altered soil conditions and disturbance regimes of intensive agriculture [@Tasser2002-om].
